---
title: "Segmented control charts with segqcc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Segmented control charts with segqcc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      fig.width = 8, fig.height = 4.5, fig.align = "center")
library(segqcc)
```

## The problem

A Shewhart chart assumes one process: a single centre, a single pair of control
limits, estimated from the whole series. When the process trends or switches
regime that assumption fails, and the chart reports it in the least useful way
possible — a long run of points outside limits that describe a state the
process left long ago.

The bundled `segxbar` data set is a variable process observed in 400 samples of
8 units, built in three stretches with means 100, 101.5 and 99.5. Charted the
ordinary way:

```{r global, fig.cap = "One global set of limits for a process with two shifts."}
suppressPackageStartupMessages(library(qcc))
groups <- qcc.groups(segxbar$value, segxbar$sample)
invisible(qcc(groups, type = "xbar", nsigmas = 3))
```

Every sample of the middle stretch is an alarm. The chart is not wrong, but it
tells us only that the mean moved — not where, and not whether the process was
stable on either side of the move.

## The segmented chart

`segmented_qcc()` locates the change points in the charted statistic and
recomputes the centre and the limits inside each segment:

```{r segmented}
fit <- segmented_qcc(segxbar$value, type = "xbar", sample = segxbar$sample)
fit
```

The two change points sit exactly where the process was built to move, and each
stretch is now judged against the limits that applied to it:

```{r segmented-plot, fig.cap = "Per-segment limits, drawn only across their own segment."}
plot(fit)
```

One sample is out of control — the sort of signal the chart is for, and which
the global chart buried under 120 false ones.

The result is an ordinary `qcc` object carrying per-sample limits, with three
extra components:

```{r components}
fit$change.points
fit$segments
head(fit$out_of_control)
```

## Chart types

The same pipeline works for every chart `qcc` provides. Attribute charts take
one value per sample, so no `sample` argument:

```{r attribute}
segmented_qcc(segc$value, type = "c")
```

`p` and `np` need `sizes`, and `u` needs `area`:

```{r p-chart}
segmented_qcc(segp$value, type = "p", sizes = segp$sizes)$segments
```

When the process is observed one unit at a time there is no within-sample
variability to estimate sigma from, and the chart has to be the individuals
one. Use `type = "xbar.one"`:

```{r individuals}
segmented_qcc(segind$value, type = "xbar.one")$segments
```

Asking for `"xbar"` on samples of size 1 is an error rather than a silent
fallback, because the two charts estimate sigma from different things:

```{r xbar-one-error, error = TRUE}
segmented_qcc(segind$value, type = "xbar", sample = seq_len(nrow(segind)))
```

For grouped samples, the R chart is available for sizes up to 25 — the range
over which `qcc` tabulates its d2/d3 constants — and the S chart for any size.

## Why the statistic is standardised

`changepoint`'s normal-likelihood cost assumes a unit-variance series. A
control-chart statistic almost never has one, and the mismatch does not degrade
the detector gently: it over-segments in proportion to the dispersion of the
series. On a homogeneous series — no change points at all — `cpt.mean` alone
finds plenty:

```{r scale-demo}
suppressPackageStartupMessages(library(changepoint))
set.seed(1)
sapply(c(1, 2, 5, 20), function(s)
  length(cpts(cpt.mean(rnorm(400, 100, s), method = "PELT", penalty = "MBIC"))))
```

`segmented_qcc()` therefore standardises the statistic before handing it to the
detector, by default with the moving-range estimate of sigma. Unlike `sd()`,
that estimate is not inflated by the very level shifts being looked for:

```{r scale-fixed}
set.seed(1)
sapply(c(1, 2, 5, 20), function(s)
  length(segmented_qcc(rnorm(4000, 100, s), type = "xbar",
                       sample = rep(1:400, each = 10))$change.points))
```

Detection is now invariant to the units of the data, which is what lets the
same call be pointed at an unfamiliar series. Change points are positions, so
nothing has to be transformed back; the limits are always recomputed from the
original data. `scale = "none"` restores the unstandardised behaviour for
anyone who wants it.

## Series that cannot be segmented

Pointed at arbitrary data, the function will meet series it cannot split. None
of these is an error: you get an ordinary single-segment control chart, a
warning, and the reason in `$notes`.

```{r degenerate}
fit <- segmented_qcc(c(3, 5, 4, 6), type = "c", min_seg_len = 3)
fit$notes
nrow(fit$segments)
```

The same applies one segment at a time. An individuals chart needs two samples
for a moving range, so a one-sample segment has no limits of its own and
borrows those of the whole series, flagged in `limits_from`:

```{r fallback}
set.seed(7)
v <- c(rnorm(30, 50, 1), 80, rnorm(30, 50, 1))
segmented_qcc(v, type = "xbar.one")$segments
```

## Tuning the detector

`method`, `penalty` and `min_seg_len` are passed to `changepoint`; `cpt_stat`
chooses whether to look for a change in the mean, the variance or both. A
stricter penalty yields fewer, larger segments:

```{r tuning}
for (pen in c("AIC", "BIC", "MBIC")) {
  n <- length(segmented_qcc(segc$value, type = "c", penalty = pen)$change.points)
  cat(pen, ":", n, "change point(s)\n")
}
```

## Sample labels

Samples are ordered by first appearance, so labels can be anything — integers
with gaps, characters, factors — and the time order of the data is never
silently permuted:

```{r labels}
set.seed(31)
v   <- rnorm(600, rep(c(100, 106), each = 300), 1)
lab <- rep(sprintf("batch-%02d", c(30:1, 60:31)), each = 10)
segmented_qcc(v, type = "xbar", sample = lab)$change.points
```
